### 2016-07-06 case-studies.tex
edits to get to 12 pages

\section{Case Studies}
\label{sec:case-studies}

In this section, we present two case studies that demonstrate the
applicability of our framework in varied settings, and its consequent
practical utility. Both our case studies target Ruby-on-Rails
applications. To analyze these, and other similar open-source
applications, we have developed a symbolic execution engine that
compiles a Rails program to an abstract program in an extended version
of \txnimp, along with some proof automation support.\footnote{Details
   can be found in the supplementary material.  Source code for our
  implementation, along with applications and experiments, are
  available at the following anonymized url:
  \url{https://github.com/popl17/popl17-code}} The focus of
our formal verification efforts are such translated \txnimp programs.

@@

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing store consistency levels, we
re-implemented the counter as an op-based \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels (both on-demand; {\sc ec} is default), and ANSI SQL
isolation levels. Keeping the consistency level at {\sc ec}, we then
repeated our experiments with committed reads and serializable
increments - a combination which was proved to preserve monotonicity
on an {\sc sc} store, but nonetheless observed monotonicity
violations. The reason for such violations became evident when
we instantiated our proof framework for an {\sc ec} store.
Under this instantiation, it is impossible to derive the
\C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
that successive reads in a session witness monotonically increasing
state. However, assuming {\sc cc} store consistency for reads allowed
us to complete the proof, and recover the monotonicity invariant. In
fact, we were able to complete the proof even after relaxing the
serializability requirement for increments, just by changing the
definition of the $\interp{S}(X)$ (\S\ref{sec:opsem}) to return
the value of the largest write-to-$X$ instead of the latest
write-to-$X$ in $S$.  In other words, if the reads chose the largest
valued write effect, instead of the latest write effect they witness,
then counter increments need not be serializable for reads to witness
a monotonically increasing count.\footnote{The trade-off is that the
view count is now no longer accurate, although it grows
monotonically.} We implemented this version of the counter as an
op-based RDT, which uses {\sc cc} reads and {\sc ec} increments, both
un-isolated (i.e., \emph{Read Uncommitted}), and obtained a 52\%
improvement in overall latency compared to the version that uses
serializable increments.
\vspace*{-3pt}

\paragraph{Microblog.} Our second set of experiments focused on a
Twitter-like microblogging application taken from~\cite{railsbook}.
An important invariant that must be preserved is referential integrity
between microposts (tweets) and authors. The application enforces this
invariant entirely in user space by relying on database-backed
transactions (this behaviour is typical of Rails
applications~\cite{bailisferal}).  For instance, the deletion of a
user is carried out in a transaction that also includes the deletion
of microposts authored by the user.  However, it is possible for
concurrent observers to nonetheless witness invariant violations. To
quantify the likelihood of such anomalies, we first populated a
Postgres database with 1000 user accounts, in which each user is
associated with 50 microposts and 20 followers, chosen uniformly at
random. We then constructed experiments with 64 concurrent clients
performing deletions and timeline reads of randomly chosen users in a
1:7 ratio, and witnessed 250 violations of referential integrity (the
isolation level was left at the default {\sc rc} level for these
experiments). To determine appropriate (correct) isolation levels, we
then instantiated our framework for an {\sc sc} store with various
ANSI SQL isolation levels for user deletions and timeline reads, and
obtained a proof that {\sc rc} is sufficient for deletions, whereas
reads require {\sc rr}. The proof is immediately applicable for MySQL,
which implements ANSI SQL isolation levels. Repeating our experiments
for MySQL with this configuration resulted in no violations (as
expected), while yielding a 38\% reduction in latency than an {\sc
  ser} configuration.  Since Postgres' isolation levels are stronger
than the corresponding ANSI levels, \rulelabel{RG-Conseq} rule
(Fig.~\ref{fig:rg-rules}) lets us immediately deduce that the same
choice of isolation levels also work on Postgres.

>>>>

\vspace*{-4pt}
\section{Case Studies}
\label{sec:case-studies}

We present two case studies below that demonstrate the applicability
of our framework and its practical utility. Both examples
target Ruby-on-Rails applications, which are commonly used to
interface applications to database systems. To analyze these, and
other similar open-source applications, we have developed a symbolic
execution engine that compiles a Rails program to an abstract program
in an extended version of \txnimp, along with some proof automation
support.\footnote{Details can be found in the supplementary material.
  Source code for our implementation, along with applications and
  experiments, are available at the following anonymized url:
  \url{...}} The focus of our formal verification efforts are such
translated \txnimp programs.

@@

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing store consistency levels, we
re-implemented the counter as an op-based \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels (both on-demand; {\sc ec} is default), and ANSI SQL
isolation levels. Keeping the consistency level at {\sc ec}, we then
repeated our experiments with committed reads and serializable
increments - a combination which was proved to preserve monotonicity
on an {\sc sc} store, but nonetheless observed monotonicity
violations.  Under this instantiation, it is impossible to derive the
\C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
that successive reads in a session witness monotonically increasing
state. However, assuming {\sc cc} for reads allowed
us to complete the proof, and recover the monotonicity invariant. In
fact, we were able to complete the proof even after relaxing the
serializability requirement for increments, just by changing the
definition of the $\interp{S}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$
in $S$.  Thus, if the reads chose the largest-valued write
effect, instead of the latest write effect they witness, then counter
increments need not be serializable for reads to witness a
monotonically increasing count.\footnote{The trade-off is that the
  view count is now no longer accurate, although it grows
  monotonically.} We implemented this version of the counter as an
op-based RDT, which uses {\sc cc} reads and {\sc ec} increments, both
un-isolated (i.e., \emph{Read Uncommitted}), and obtained a 52\%
improvement in overall latency compared to the serializable original.
\vspace*{-3pt}

\paragraph{Microblog.} Our second set of experiments focused on a
Twitter-like microblogging application taken from~\cite{railsbook}.
An important invariant that must be preserved is referential integrity
between microposts (tweets) and authors. The application enforces this
invariant entirely in user space by relying on database-backed
transactions (this behaviour is typical of Rails
applications~\cite{bailisferal}).  For instance, the deletion of a
user is carried out in a transaction that also includes the deletion
of microposts authored by the user.  However, it is possible for
concurrent observers to nonetheless witness invariant violations. To
quantify the likelihood of such anomalies, we first populated a
Postgres database with 1000 user accounts, in which each user is
associated with 50 microposts and 20 followers, chosen uniformly at
random. We then constructed experiments with 64 concurrent clients
performing deletions and timeline reads of randomly chosen users in a
1:7 ratio, and witnessed 250 violations of referential integrity (the
isolation level was left at the default {\sc rc} level for these
experiments). To determine appropriate (correct) isolation levels, we
then instantiated our framework for an {\sc sc} store with various
ANSI SQL isolation levels for user deletions and timeline reads, and
obtained a proof that {\sc rc} is sufficient for deletions, whereas
reads require {\sc rr}. The proof is immediately applicable for MySQL,
which implements ANSI SQL isolation levels. Repeating our experiments
for MySQL with this configuration resulted in no violations (as
expected), while yielding a 38\% reduction in latency compared to an
{\sc ser} configuration.  Since Postgres' isolation levels are
stronger than the corresponding ANSI levels, the \rulelabel{RG-Conseq}
rule (Fig.~\ref{fig:rg-rules}) lets us immediately deduce that the
same choice of isolation levels also work on Postgres.

### 2016-07-06 store-consistency.tex
edits to get to 12 pages

Reasoning under weakly-consistent replication has received special
attention in recent work~\cite{gotsmanpopl16}. Our operational
semantics and proof system are general enough to admit replication as
a special case of our formulation. In this section, we explain how the
standard artifacts of weakly-consistent replication manifest in our
reasoning framework.

@@

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. Both withdraw transactions are run at {\sc si}
isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd2)}}$. We assume an {\sc sc} store, such as a
conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
the maximul visibility weakening (Def.~\ref{def:maxvis}) of {\sc sc}
w.r.t. $\I_c$, as described in \S\ref{sec:store-consistency}. In
concrete terms, this means that an effect added to $\E.\A$ by the
current thread or a concurrent thread immediately becomes visible to
the subsequent operations unless we produce a witness that such
visibility violates $\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note
that the language of our specifications is a decidable subset of the
first-order logic. In particular, our specifications and
rely-guarantee rules are carefully crafted to avoid transitive
closures and quantification over infinite domains. The domain of
assertions is often linear arithmetic, which is also decidable.
Consequently, the proof obligations generated by the application of
our rules can often be discharged via an SMT solver. We nonetheless
give full details in the following.

@@

  \item The value written to \C{B} is either $\C{k-a1-a2}$ or
  $\C{k-a1}$ depending on whether or not \C{Wd1} has already
  committed. This is clearly necessary to deduce the value of \C{B}
  before and after \C{Wd1} commits.
\end{itemize}
The resultant rely relation ($R_1$) is shown below (note: we
write $(\E'-\E) \Vdash {\committed(\C{Wd2})}$ to express that \C{Wd2}
is not committed in $\E$, but committed in $\E'$.):

\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd1})} \conj
        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
%       \neg\underE{\committed(\C{Wd2})} \conj \I(\E) \conj \I(\E')\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{B} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{B=k-a1-a2} \\
%   & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
%       (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
    & & \hspace*{0.8in}\Rightarrow \C{B=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{B} \\
    & & \hspace*{0.37in} \conj \underE{\committed(\C{Wd1})} \conj
        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
%       \C{COMMIT(Wd2)} \in (\E'.\A-\E.\A) \\
    & & \hspace*{0.6in}\Rightarrow \C{B=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{B} \\
    & & \hspace*{0.37in}\conj \forall(\eta \in \E'.\A - 
        \E.\A).~\txn(\eta) = \C{Wd2}\}\\
\end{array}
\end{smathpar}
For the sake of completeness, we also explicitly represent the fact
that all the effects added by $R_1$ are of \C{Wd2}. This is needed to
prove that an interference from \C{Wd2} does not affect the local
state of \C{Wd1}. $R_1$ refers to labels (\C{Wd1} and \C{Wd2}) of both
the transactions. This is allowed because labels are artifacts of the
execution $\E$; they are not (local) variables.

The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
$R_1$ does not commit \C{Wd2}, then the value of \C{B} is unaffected.
If it commits \C{Wd2}, then since \C{Wd1} had already performed (read)
operations that cannot witness \C{Wd2}, we have
$\C{Wd2}\not\visar\C{Wd1}$, and since $\phi_4$ only makes assertions
under the case $\C{Wd2}\visar\C{Wd1}$, it is stable. For $\phi_5$,
{\sc si} condition should be used to show that any interference from
\C{Wd2} is invalid. The proof proceeds on the same lines as the Hoare
proof discussed above; it is also discussed in detail in
\S\ref{sec:motivation}.

The final aspect of the proof is to show that any interference from
\C{Wd1} is contained in its guarantee relation $G_1$, which is also
the rely relation $R_2$ for \C{Wd2}. Since the implementation of
\C{Wd2} is same as \C{Wd1}, $R_2$ (hence $G_1$) is same as $R_1$ shown
above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our focus
on the assignment statement, the guarantee proof requires us to show
that if executing the assignment takes the execution from $\E$ to
$\E'$, such that $\E'$ satisfies $\I$, then $(\E,\E') \in G_1$.  The
proof follows from the \rulelabel{RG-Asgn} rule, which establishes
that (a). the assignment statement does not commit \C{Wd2}, and (b).
it adds a write effect $\eta$ such that $\txn(\eta)=\C{Wd1}$. This
finishes the proof for the triple in focus. The proofs for other
triples of \C{Wd1} can also be constructed similarly. The combination
of correctness proofs for \C{Wd1} and \C{Wd2} yields a correctness
proof for the program.

>>>>

Reasoning under weakly-consistent replication has received special
attention in recent work~\cite{gotsmanpopl16}. Our operational
semantics and proof system are general enough to admit replication as
a special case of our formulation, as we explain below.

@@

We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
in greater detail. Both withdraw transactions are run at {\sc si}
isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
\underE{\C{SI(Wd2)}}$. We assume an {\sc sc} store, such as a
conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
the maximal visibility weakening (Def.~\ref{def:maxvis}) of {\sc sc}
w.r.t. $\I_c$, as described in \S\ref{sec:store-consistency}. In
concrete terms, this means that an effect added to $\E.\A$ by the
current thread or a concurrent thread immediately becomes visible to
the subsequent operations unless we produce a witness that such
visibility violates $\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note
that the language of our specifications is a decidable subset of the
first-order logic. In particular, our specifications and
rely-guarantee rules are carefully crafted to avoid transitive
closures and quantification over infinite domains. The domain of
assertions is drawn from linear arithmetic, which is also decidable.
Consequently, the proof obligations generated by the application of
our rules can often be discharged via an SMT solver.

@@

  \item The value written to \C{B} is either $\C{k-a1-a2}$ or
  $\C{k-a1}$ depending on whether or not \C{Wd1} has already
  committed. This is clearly necessary to deduce the value of \C{B}
  before and after \C{Wd1} commits.
\end{itemize}
The resultant rely relation ($R_1$) is shown below (note: we
write $(\E'-\E) \Vdash {\committed(\C{Wd2})}$ to express that \C{Wd2}
is not committed in $\E$, but committed in $\E'$.):
\vspace*{-1.5pt}
\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd1})} \conj
        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
%       \neg\underE{\committed(\C{Wd2})} \conj \I(\E) \conj \I(\E')\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{B} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{B=k-a1-a2} \\
%   & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
%       (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
    & & \hspace*{0.8in}\Rightarrow \C{B=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{B} \\
    & & \hspace*{0.37in} \conj \underE{\committed(\C{Wd1})} \conj
        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
%       \C{COMMIT(Wd2)} \in (\E'.\A-\E.\A) \\
    & & \hspace*{0.6in}\Rightarrow \C{B=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{B} \\
    & & \hspace*{0.37in}\conj \forall(\eta \in \E'.\A - 
        \E.\A).~\txn(\eta) = \C{Wd2}\}\\
\end{array}
\end{smathpar}

\noindent For the sake of completeness, we also explicitly represent the fact
that all the effects added by $R_1$ are from \C{Wd2}. This is needed to
prove that an interference from \C{Wd2} does not affect the local
state of \C{Wd1}.

The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
$R_1$ does not commit \C{Wd2}, then the value of \C{B} is unaffected.
If it commits \C{Wd2}, then since \C{Wd1} had already performed (read)
operations that cannot witness \C{Wd2}, we have
$\C{Wd2}\not\visar\C{Wd1}$, and since $\phi_4$ only makes assertions
under the case $\C{Wd2}\visar\C{Wd1}$, it is stable. For $\phi_5$, the
{\sc si} condition is used to show that any interference from \C{Wd2}
is invalid. The proof proceeds on the same lines as the proof
discussed above, and is also discussed in detail
in\S\ref{sec:motivation}.

The final aspect of the proof is to show that any interference from
\C{Wd1} is contained in its guarantee relation $G_1$, which is also
the rely relation $R_2$ for \C{Wd2}. Since the implementation of
\C{Wd2} is same as \C{Wd1}'s, $R_2$ (hence $G_1$) is same as $R_1$
shown above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our
focus on the assignment statement, the guarantee proof requires us to
show that if executing the assignment takes the execution from $\E$ to
$\E'$, such that $\E'$ satisfies $\I$, then $(\E,\E') \in G_1$.  The
proof follows from the \rulelabel{RG-Asgn} rule, which establishes
that (a). the assignment statement does not commit \C{Wd2}, and (b).
it adds a write effect $\eta$ such that $\txn(\eta)=\C{Wd1}$.  The
proofs for other triples of \C{Wd1} can also be constructed
similarly. The combination of correctness proofs for \C{Wd1} and
\C{Wd2} yields a correctness proof for the program.
